---
title: "gapminder"
author: "OM"
date: "March 13, 2018"
output: html_document
---

```{r}
install.packages("tidyverse")
```

```{r}
library(tidyverse)

##Learning tidyr and dplyr with gapminder data

```{r}
gapminder <- read_csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
```


```{r}
dim(gapminder)
ncol(gapminder)
nrow(gapminder)
```

```{r}
str(gapminder)
```


```{r}
unique(gapminder$country)
```

```{r}
unique(gapminder[,"continent"])
unique(gapminder$continent)
```

```{r}
summary(gapminder)
```

```{r}
install.packages("skimr")
```
```{r}
library(skimr)
```
```{r}
skim(gapminder)
```

##Filtering

```{r}
#filter()
#same since we have dplyr loaded as below - and can also specify explicitly -
dplyr::filter(gapminder, country == "Switzerland")
# stats::filter() is masked by dplyr's filter function but could be called
```
```{r}
?filter
```
```{r}
filter(gapminder, lifeExp<30)
```


```{r}
filter(gapminder, country == "Afghanistan", year == max(year))
filter(gapminder, country == "Rwanda", year == max(year))
```
```{r}
dplyr::select(gapminder, country, year, pop)
```

```{r}
gapminder_no_cont <- dplyr::select(gapminder, -continent)
```
```{r}
dplyr::filter(gapminder_no_cont, year == 2007)
```
##Creating pipelines with %>%
Instead of creating a pile of intermediate variables, can pipe commands together 

```{r}
select(gapminder, -continent) %>% filter(year == 2007)
```
This injects the output of one operation into the next operation - allows one to make the code very readable and stepwise. This is a strength of dplyr since this function is not in base R. Actuall lives in the package migrateR which is included in dplyr.

```{r}
#mutate(gapminder) or alternatively (and non-sci notation display)
options(scipen=10000)
gapminder %>%
  mutate(gdp = gdpPercap*pop) %>%
  arrange(desc(gdp))


```
##Grouping

```{r}
options(scipen=10000)
gapminder %>%
  group_by(country) %>%
  mutate(gdp = gdpPercap*pop, max_gdp = max(gdp)) %>%
  ungroup()
#good practice to ungroup after a manipulation
```
```{r}
options(scipen=10000)
gapminder %>%
  group_by(country) %>%
  mutate(gdp = pop*gdpPercap) %>%
  summarize(max_gdp = max(gdp)) %>%
  arrange(desc(max_gdp))
```
##Combining Datasets
Need a column of information in common, and need unique IDs for rows in both datasets
Left join: left table has priority, join matching rows left, right, byfield
Right join: right table has priority, join matching rows, left, right, byfield
Inner join: retain only rows in both sets, join rows, left, right, byfield
Full join: retains all values, all rows, left, right, byfield

In this example: have CO2 data for selected countries for 2007.

```{r}
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
co2
```
Good idea to keep all variables and names lower case if possible.
Combine with gapminder data.

```{r}
gap_2007 <- gapminder %>%
  filter(year == 2007)
#check to see if number of countries match
length(unique(gapminder$country))
```
Yes - matches number of countries n 2007. 
Can often look at fails to see discrepancies; stringr program helps remap names; use unique IDs when available - better than country names in this example - and use authoritative sources where possible (e.g., UN for country names)

```{r}
left_join(gap_2007, co2, by = "country")
```
```{r}
right_join(gap_2007, co2, by = "country")
```
##setdiff
Can also explicitly test for differences before joining to test for misspellings, potential sources of inadvertent errors.

```{r}
setdiff(co2$country, gap_2007$country)
```








##Tidyr
```{r}
AirPassengers
```

In tidyr can gather, spread, separate, unite.
In the AirPassengers dataset would separate into 3 columns: year, month, number

###Tidying gapminder data from wide format

```{r}
gap_wide <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")
gap_wide

```


```{r}
library(dplyr)
library(tidyr)

gap_long <- gap_wide %>%
  gather(key   = obstype_year,
         value = obs_values)
```

```{r}
gap_long <- gap_wide %>%
  gather(observation, observ_value, -continent, -country)

```

```{r}
gap_better <- gap_wide %>%
  gather(observation, observ_value, -continent, -country) %>%
  separate(observation, into = c("obs_type", "year"), sep = "_", convert = TRUE)

gap_better

```

```{r}
unique(gap_better$obs_type)

```

Note: errors earlier were caused by restarting R - had to reload libraries (dplyr and tidyr) - clue was not recognizing pipeline %>%

###tidyr functions: gather, spread, separate, unite

To make a ollaborative branch rename the master branch gh-ages